### 2021-02-02 conclusions.tex
intial commit

\section{Conclusions and Future Work}
We have demonstrated\dots

\paragraph{Future Work}
Developments demonstrated in this paper open up exciting perspectives
for future research.
\begin{itemize}
  \item Do something with the semantics preservation.  We need to
    define semantics for Agda's internal syntax (or the subset
    that is being used to embed the DSL).  Then define the mapping
    from the target language syntax into Agda's internal syntax.
    Prove that extraction preserves the semantics for all the valid
    inputs.  In this case the only point of failure would be the
    mapping, which is much smaller than the entire extractor.
  \item Consider irrelevance annotations to function arguments.
    Agda makes it possible to annotate function arguments as irrelevant,
    meaning that it is guaranteed that they are only used during
    typechecking and may be erased at runtime.  Right now such annotations
    are ignored, but such erasures would benefit performance.
  \item Support let constructs in the internal syntax.  Currently
    let-s in Agda are treated purely as a syntactic sugar.  After
    desugaring all the lets are inlined, which makes the terms
    very large, and has a negative impact on performance when
    being executed without any optimisations.  Adding lets to the
    internal syntax would solve this problem, unfortunately this is
    far from trivial, as such a change generates enormous changes
    in the typechecker.
  \item Elegant extraction of recursive functions defined via
    well-founded recursion.
\end{itemize}

### 2021-03-02 conclusions.tex
New introduction.
Conclusions.

@@
-\section{Conclusions and Future Work}
-We have demonstrated\dots
+\section{\label{sec:concl}Conclusions and Future Work}
 
-\paragraph{Future Work}
-Developments demonstrated in this paper open up exciting perspectives
-for future research.
-\begin{itemize}
-  \item Do something with the semantics preservation.  We need to
-    define semantics for Agda's internal syntax (or the subset
-    that is being used to embed the DSL).  Then define the mapping
-    from the target language syntax into Agda's internal syntax.
-    Prove that extraction preserves the semantics for all the valid
-    inputs.  In this case the only point of failure would be the
-    mapping, which is much smaller than the entire extractor.
-  \item Consider irrelevance annotations to function arguments.
-    Agda makes it possible to annotate function arguments as irrelevant,
-    meaning that it is guaranteed that they are only used during
-    typechecking and may be erased at runtime.  Right now such annotations
-    are ignored, but such erasures would benefit performance.
-  \item Support let constructs in the internal syntax.  Currently
-    let-s in Agda are treated purely as a syntactic sugar.  After
-    desugaring all the lets are inlined, which makes the terms
-    very large, and has a negative impact on performance when
-    being executed without any optimisations.  Adding lets to the
-    internal syntax would solve this problem, unfortunately this is
-    far from trivial, as such a change generates enormous changes
-    in the typechecker.
-  \item Elegant extraction of recursive functions defined via
-    well-founded recursion.
-\end{itemize}
+In this paper we investigate the notion of extraction.  This
+is the ability to express programs hand-in-hand with custom
+code generators for them.  We have demonstrated that
+compile-time metaprogramming is sufficient to equip
+a programming language with extraction capabilities.
+A well-known conundrum of choosing between deep and
+shallow embedding can be resolved if the host language
+is extraction-capable.  While enjoying the benefits
+of shallow embedding, there is a full access to the internal
+structure of the embedded programs.
+
+We use this idea in the context of dependently-typed
+languages to create verified implementations that can
+be used in the context of an existing programming language.
+We embed a language of interest in a theorem-prover, using
+dependent types to encode the properties of interest.
+Using extraction we bring the verified implementation
+back into the original language.
+
+We have demonstrated the approach by implementing
+three embedded languages and two extractors using
+Agda as our host.  We have demonstrated the
+changes to Agda that we had to do on the way, and we
+used our embedding to implement (and verify) a practical
+application --- a convolutional neural network.
+
+The main advantages are twofold.  First,
+our solution is fine-grain --- we can chose what part
+of the application to embed, and what constructs of the
+host language to extract.  Secondly, extractors
+can use the full power of dependent types guaranteeing
+some safety properties of interest.
+
+In order to make the proposed idea even more powerful,
+we would require improvements to metaprogramming capabilities
+of theorem provers.
+
+Most importantly, right now it is not guaranteed that the
+extracted code preservers the semantics of the original
+implementation.  Essentially, one has to trust that the
+extractor is doing the right job.  While we rarely see
+fully-verified compiler backends in the real world, in
+the proposed environment we are very close to enabling this.
+We would need a formal semantics of the reflected language
+and the proof that reflected programs respect it.
+While this is non-trivial, a system like Agda could do
+this in principle.
+
+There is a number of improvements that can be added to
+Agda and our extractors to make the resulting code more
+efficient.  Supporting `let's in the internal syntax
+would help to preserve sharing.  Recognising irrelevance
+annotations in the extractors would help to eliminate
+unused function arguments.  Introducing proper language
+primitives to specify what exactly is an embedding would
+be helpful.
+
+Overall, this work only scratches the surface of extraction-based
+compilation.  We never considered alternative theories supported
+by theorem provers, \eg{} cubical type theory in Agda; we did not
+consider recursive metaprogramming; we did not consider integrating
+optimisations of extracted programs besides of what rewriting rules
+are capable to do.  All of these offer exciting research opportunities
+on the way to make verified software easily accessible.
+To paraphrase Jim Morrison: ``Safety and no surprise, The End''.

### 2021-03-02 conclusions.tex
Related work.
Shrink/edit pass on conclusion (under 25 pages!!!)

--- 2021-02-02 conclusions.tex
@@
-\section{Conclusions and Future Work}
-We have demonstrated\dots
+\section{\label{sec:concl}Conclusions and Future Work}
 
-\paragraph{Future Work}
-Developments demonstrated in this paper open up exciting perspectives
-for future research.
-\begin{itemize}
-  \item Do something with the semantics preservation.  We need to
-    define semantics for Agda's internal syntax (or the subset
-    that is being used to embed the DSL).  Then define the mapping
-    from the target language syntax into Agda's internal syntax.
-    Prove that extraction preserves the semantics for all the valid
-    inputs.  In this case the only point of failure would be the
-    mapping, which is much smaller than the entire extractor.
-  \item Consider irrelevance annotations to function arguments.
-    Agda makes it possible to annotate function arguments as irrelevant,
-    meaning that it is guaranteed that they are only used during
-    typechecking and may be erased at runtime.  Right now such annotations
-    are ignored, but such erasures would benefit performance.
-  \item Support let constructs in the internal syntax.  Currently
-    let-s in Agda are treated purely as a syntactic sugar.  After
-    desugaring all the lets are inlined, which makes the terms
-    very large, and has a negative impact on performance when
-    being executed without any optimisations.  Adding lets to the
-    internal syntax would solve this problem, unfortunately this is
-    far from trivial, as such a change generates enormous changes
-    in the typechecker.
-  \item Elegant extraction of recursive functions defined via
-    well-founded recursion.
-\end{itemize}
+In this paper we investigate the idea of developing
+embedded programs hand-in-hand with custom
+code generators for them. We solve the well-known
+conundrum of choosing between deep and shallow
+embedding by leveraging the power of
+reflection.
+This allows us to enjoying the benefits
+of shallow embedding, while keeping full access to the internal
+structure of the embedded programs.
+
+We apply this idea in the context of dependently-typed
+languages to create verified implementations that can
+be used in the context of an existing programming language.
+We embed the target language in a theorem-prover, using
+dependent types to encode the properties of interest.
+Using reflection we bring the verified implementation
+back into the original language.
+
+We have demonstrated the approach by implementing
+three embedded languages and two extractors, using
+Agda as our host. Along the way we made some
+improvements to the reflection API of Agda, and in the end we
+used our embedding to implement (and verify) a practical
+application --- a convolutional neural network.
+
+The main advantages of our approach are twofold.  First,
+our solution is fine-grained --- we can chose what part
+of the application to embed, and what constructs of the
+host language to extract.  Secondly, our extractors
+can use the full power of dependent types to guarantee
+safety of our embedded programs.
+
+Right now we cannot yet guarantee that the
+extracted code preservers the semantics of the original
+implementation. While we rarely see
+fully-verified compiler backends in the real world,
+our approach is very close to enabling this.
+We would need a formal semantics of the reflected language
+and the proof that reflected programs respect it.
+While this is non-trivial, a system like Agda could do
+this in principle.
+
+There is a number of improvements that can be added to
+Agda and our extractors to make the resulting code more
+efficient.  Supporting \AK{let}s in the internal syntax
+would help to preserve sharing.  Recognising irrelevance
+annotations in the extractors would help to eliminate
+unused function arguments.  Introducing proper language
+primitives to specify what exactly is an embedding would
+be helpful. And finally, having access to more of
+Agda's internals such as case trees would help to
+generate more performant code.
+
+Overall, this work only scratches the surface of extraction-based
+compilation.  We never considered alternative theories supported
+by theorem provers, \eg{} cubical type theory in Agda; we did not
+consider recursive metaprogramming; we did not consider integrating
+optimisations of extracted programs other than what rewriting rules
+are capable to do.  All of these offer exciting research opportunities
+on the way to make verified software easily accessible.
+To paraphrase Jim Morrison: ``Safety and no surprise, The End''.

### 2021-04-23 review-response.md
Starting to put response together.

# Structure of the response

We would like to thank all the reviewers for their time and helpful comments!
We structure our response in two parts.  First, we address common concerns raised
by multiple reviewers; after that, we respond to the remaining individual
comments and questions.

# Part1: Common concerns

## The main goal of the paper as we see it

The main point of the paper is to demonstrate that it is possible to define
a practically usable *dependently-typed DSL* without the necessity to
implement a typechecker, as you share one with Agda.  Extraction gives
us a way to define an arbitrary target language that we translate our DSL
for further execution.  Two practical scenarios are possible: 1) I have a
(dep.typed) DSL and I want to translate it to target X; or 2) I have a
a language X which I would like to extend with dependent types.  Both
scenarios are possible with the presented work, and to our knowledge
this was not tried before.

## General principles and correctness

As our main focus is to provide all the means to write an extractor,
the question of correctness (in some sense) is orthogonal ---
one is free to make it as correct or incorrect as one pleases.

So far we can state that, as long as you trust that Agda type system
is sound:
  1. an embedded language has a sound type system (as it is a subset
     of Agda);
  2. programs in the embedded languages have expected properties
     of progress and preservation (as they can be executed within Agda)
  3. given that the extractor preserves the dynamic semantics of Agda:
     - static semantics is encoded using assertions;
     - if an entire program is extracted from Agda, these assertions
       shall never be triggered at runtime;
     - when extracting partial programs, assertions can only fire at
       the boundaries with unsafe code.

Ideally, we would want to write a theorem in Agda that the given
extractor preserves the semantics.  For that one needs to relate the
semantics of reflected terms with the semantics of the language we
are extracting to.  Unfortunately, at the time of writing there is
no official semantics for Agda reflected terms.  While in principle
it is possible to come up with one manually, we believe that reflection
system should give us the one we could actually trust.  This is our
future work.

## Shallow embedding and its semantics

One challenge with a shallowly embedded language is how to formally define
which terms of the host language actually form the embedding.  While we do
not have any syntactical means to do so, as we state in line 315, any Agda
term that is recognised by the extractor is considered to be the part of
the embedded language.  In that sense, semantics of the embedding is always
given by the semantics of Agda, and the language itself is defined by the
extractor.

## Pull Requests inclusion in the paper

We consider the changes that we made to Agda essential to the work that we
present in the paper.  As our approach has not been tried before, obviously
some required functionality is lacking, so we fixed it.  As for inclusion
in the paper, unfortunately we didn't find a better way of presenting these,
and according to our read of the definition of "lightweight double-blind"
inclusion of the links seem to be acceptable as we do not expose our names
or institutions, and as the work has been done specifically for this paper
we did not have to address it in third person.

We summarise the changes that we did to Agda's reflection system:
  - allow accessing the actual typing context (telescope) of the given term;
  - selective reduction and normalisation of the terms;
  - access typing information of the implicit arguments after normalisation
    -- the most challenging change.

# Part2: Remaining concerns

-- XXX stopped here XXX ---

ICFP 2021 Paper #150 Reviews and Comments
===========================================================================
Paper #150 Choosing is Losing: How to combine the benefits of shallow and
deep embeddings through reflection

Review #150A
===========================================================================

Overall merit
-------------
C. I would not accept this paper but will not argue strongly against
   accepting it.

Reviewer expertise
------------------
X. Expert

Paper summary
-------------
1 Introduction: the idea of developing embedded programs hand-in-hand with custom
code generators for them.
2 Background, mostly standard stuff on Agda.  A little on quote/unquote.
3 Basic extraction, illustrated using Kaleidoscope, a minimalist 1st order language from the LLVM team.  Shows how to extract a Kaleidoscope program from an Agda term.  Example of the binary logarithm. Use of Agda rewrite rules.
4 Embedding the array language SaC in Agda, and extracting to it.
A key progression compared to 3 is that SaC has a sophisticated array type system, that can be represented (more or less) within Agda.
5 Embeds a form of APL within Agda.  The embedding is not all of APL; one key concept retained is automatic casting of scalars to vectors and arrays for use with binary operators.

Pros and Cons
+ new idea to embed DSL as a shallow embedding in Agda, but use reflection to access the syntax trees of the shallow embedding, a kind of deep embedding
+ the shallow embedding allows properties to be type-checking, while reflection allows extraction of the verified program back to the original language
+ three separate language examples: core language Kaleidoscope, array language SaC, and APL
+ dependently typed array operations simply fall out of the encoding in the underlying system
- no discussion of performance, eg, of the CNN
- the general principles here are not expressed crisply, eg, as theorems

Comments for author
-------------------
Thanks for all your work on this paper.  I learnt from reading it.

Unfortunately, the main weakness to me is that there is no general theory or crisp principles here, although that may be a weakness in the presentation, rather than the underlying coding.

For the author response, please respond to the next two paragraphs especially.

Your key contribution is the idea of a "reflection-based extractor" for a "shallowly-embeddable language".  What is the reflection-based extractor for your Kaleidoscope example?  What are its key properties?  Can they be expressed as theorems within (or possibly outside) Agda?

Another key contribution is a set of changes to Agda, described in three sections of the paper: 3.4, 3.5, and 4.2, each of which refer to pull requests. Can you describe the changes in terms of general principles that could apply in other settings?

How does the performance of the code extracted from the CNN model compare to the orginal?  I think the original is in APL and the extracted code is in SaC.

It is cool you can write APL formulas directly in Agda.

Can you say anything about what's like to program in these shallow embeddings?  Are the error messages readable?  Could they be customised to the embedded language?

Since their deep versus shallow terminology is in the title, you should cite the paper that originated the distinction between deep and shallow embeddings:

Richard J. Boulton, Andrew D. Gordon, Michael J. C. Gordon, John Harrison, John Herbert, John Van Tassel: Experience with Embedding Hardware Description Languages in HOL. TPCD 1992: 129-156

Review #150B
===========================================================================

Overall merit
-------------
C. I would not accept this paper but will not argue strongly against
   accepting it.

Reviewer expertise
------------------
Z. Some familiarity

Paper summary
-------------
This paper explores some applications of using reflection to implement code
extraction to embedded languages.  The idea is to use reflections facilities to
enable the translation of (subsets of) the host language (in this case, Agda)
into an embedded language.  Such a translation might require both type and term
translation, and might also require the generated code to include dynamic checks
to reconcile the expressive dependend types of the host with those of the
target.  The paper presents several case studies, including a translation of
Agda to Kaleidoscope (a toy language included as an example from the LLVM
framework), to a simple array language, and to APL.

Comments for author
-------------------
# Review

I had a very hard time getting anything out of this paper.  The authors espouse
the benefits of combining "shallow" and "deep" embeddings, ostensibly with the
purposes of doing things like verified extraction or code transformations, but
the paper doesn't really deliver on that promise.  Moreover, I found the
presentation very confusing.  I cannot support this paper for acceptance.

# Comments

- The introduction talks about the ability to "write a verified program
  hand-in-hand with an executor for it" (line 43; maybe you meant
  "extractor"?)), but this paper doesn't seem to include any semantics for the
  languages.  The "verification results" seem to be very limited in nature --
  providing, e.g., some additional guarantees on top of APL's weak shape-based
  type system.  Moreover, since assertations that capture those stronger
  constraints are propagated into the output, I would expect there to be some
  kind of theorem stating that those assertaions are never triggered (under
  reasonable assumptions).  There aren't (as far as I can see), and explicitly
  stated correctness or verification theorems mentioned in the paper. Is
  there  some kind of correctness guarantee that you get when translating from
  APL to SaC via your infrastructure?

- I'm not sure that I would consider the encodings of Kaleidoscope or SaC in
  this paper as "shallow embeddings": my understanding is that a shallow
  embedding uses the features of the host language (in this case, Agda) to
  implement the _semantics_ of the embedded language (often using host-level
  functions to represent embedding-level variable binding, etc.).  However, I
  didn't see any semantics ascribed to Kaleidoscope, and only minimally so for
  SaC.  It looks like this code doesn't do much more than provide the abstract
  syntax for the embedded languages in question -- based on the definitions in
  this paper could you prove anything about the behavior of a program extracted
  to Kaleidoscope?  The APL embedding seems like a true embedded language in the
  sense that its semantics are implemened in Agda.

- In several places, the paper mentions limitations of Agda's reflection API
  that were changed or enhanced while doing this work (e.g. paragraphs near line
  332, 362, 829), but these are mentioned mostly in passing and not described in
  enough detail to be meanginful in any way to the reader.  Pointing to github
  pull requests is not appropriate for a scholarly paper (and isn't kosher for
  double-blind reviewing).  These anecdotes made me wonder what the point of
  this paper is, since it's apparently *not* intended to explain any of the
  technical details involved with these changes.

- The translation of reflected terms into Kaleidoscope doesn't seem to be that
  complicated, but I would have appreciated if the discussion starting at line
  557 had walked though the code of `kompile-term` a bit more explicitly.

# Details

- line 184: says that the primitives are `quote` and `unquote` and that they
  operate "as follows", but the subsequent code uses `macro` and `quote` has
  already been explained above.

- line 331: "contains any about strings" $\to$ "contains any strings"

- line 355: This definition of decidable equality is unintelligible to a reader
  with no Agda experience.  (Also, it might be good to mention earlier that it's
  common for Agda identifiers to include symbols that are usualy considered
  reserved).

- line 787:

- line 789: The "lx type is a type of valid indices..." should probably be "lx d s type is..."

- line 1208: Seems pretty late in the paper to bury the lede about not *really* doing any kind of verification.

Review #150C
===========================================================================

Overall merit
-------------
A. I will champion accepting this paper.

Reviewer expertise
------------------
Y. Knowledgeable

Paper summary
-------------
This paper attempts to answer the choice between shallow and deep embeddings
(in particular in a dependently typed context), by refusing to answer the question
in the first place. Instead, the authors rely on compile-time metaprogramming facilities
in a host language to specify a subset of the host language (in this case Agda) that
serves as the shallow-ish embedding that can be extracted to the target language of choice,
while still allowing for reasoning about its code using all the convenience of a
language like Agda. The paper then proceeds to give two extensive and convincing case
studies using this approach.

Comments for author
-------------------
This paper was _really_ fun to read. It opens with an intriguing idea - can we use
reflection facilities in a dependently typed language - to get the best out of both
worlds: those of shallow and deep embeddings. And it technically delivers that!
The 2/2.5 case studies presented demonstrate convincingly that this is a viable
approach - the matrix multiplication example in particular is an excellent
demonstration that _users_ of the embedding can enjoy strong static guarantees
in a seamless manner like in a shallow embedding. What I'm less convinced by
is the ease of use for the _implementor_. In particular, promises of the introduction
like "No need to touch internals of the host language" are true only in spirit. Yes,
in a language with ideal reflection capabilities what the authors describe would be
even more convenient than described in the paper, but even for this work the authors
had to submit multiple pull requests to change Agda internals and expose additional
capabilities. A deep embedding for Kaleidoscope for example sounds simple enough to be
much more natural to implement - but I guess then you wouldn't enjoy the ease-of-use
that we're seeing here.

Still, even though the paper reads like a tour-de-force in Agda-fu, it is really well written,
with multiple carefully selected and well explained examples. Do I think that after
reading this paper I will be reaching when implementing my next DSL for this technique
(or even that I'm able to reach for this technique without significant investment in
metaprogramming internals)? Perhaps; probably not. But it is a novel first step taking
an old dilemma in a very interesting direction, while showing concrete potential.

======== Minor ==========

- L. 27: Not the canonical citation for Coq I think. Usually people reach for the reference manual

- Pull request links should probably have been elided for the review version
(I haven't clicked them yet as of the time of this writing, but I'm assuming the break double blind)

### 2021-04-25 review-response.md
Some changes to part one of the response

@@
-# Structure of the response
-
-We would like to thank all the reviewers for their time and helpful comments!
-We structure our response in two parts.  First, we address common concerns raised
-by multiple reviewers; after that, we respond to the remaining individual
-comments and questions.
+We would like to thank the reviewers for their time and helpful comments. We
+will start this response by addressing common concerns raised by multiple
+reviewers; after that, we respond to the remaining individual comments and
+questions.
 
@@
 
-## The main goal of the paper as we see it
+## The main objective (reviewer B)
 
-The main point of the paper is to demonstrate that it is possible to define
-a practically usable *dependently-typed DSL* without the necessity to
-implement a typechecker, as you share one with Agda.  Extraction gives
-us a way to define an arbitrary target language that we translate our DSL
-for further execution.  Two practical scenarios are possible: 1) I have a
-(dep.typed) DSL and I want to translate it to target X; or 2) I have a
-a language X which I would like to extend with dependent types.  Both
-scenarios are possible with the presented work, and to our knowledge
-this was not tried before.
+Our main objective is to demonstrate how to define a *dependently-typed DSL*
+without implementing a full typechecker, as you share one with Agda. Extraction
+allows us to translate our DSL to a target language for further execution. We
+expect two possible scenarios: 1) translating an existing (dependently typed)
+DSL to target X; or 2) extend existing language X with dependent types. Both
+scenarios are possible by our approach, and are novel to the best of our
+knowledge.
 
-## General principles and correctness
+## Shallow embedding and its semantics (reviewers A and B)
 
-As our main focus is to provide all the means to write an extractor,
-the question of correctness (in some sense) is orthogonal ---
-one is free to make it as correct or incorrect as one pleases.
+For the purpose of this paper, we define any Agda term that is recognized by
+the extractor to be part of the embedded language (as noted on line 315).
+Moreover, the semantics of the embedded language is obtained by restricting the
+semantics of Agda to this subset. The writer of the extractor thus has full
+control over what parts of Agda to include or exclude.
 
-So far we can state that, as long as you trust that Agda type system
-is sound:
-  1. an embedded language has a sound type system (as it is a subset
-     of Agda);
-  2. programs in the embedded languages have expected properties
-     of progress and preservation (as they can be executed within Agda)
-  3. given that the extractor preserves the dynamic semantics of Agda:
-     - static semantics is encoded using assertions;
-     - if an entire program is extracted from Agda, these assertions
-       shall never be triggered at runtime;
-     - when extracting partial programs, assertions can only fire at
+## Lack of theory or general principles (reviewers A and B)
+
+Despite our focus on the practical parts of writing custom extractors, this
+does not mean our approach does not care about the semantics. In particular,
+the extractor maps the reflected Agda syntax of the embedded language to the
+target language, and inserts assertions corresponding to the static guarantees
+of Agda's type system. Hence, given an extractor that preserves the dynamic
+semantics of the embedded language, soundness of the Agda type system
+guarantees that none of the inserted assertions will be triggered at run-time.
+When only part of a program is extracted from Agda, assertions can only fire at
 the boundaries with unsafe code.
 
-Ideally, we would want to write a theorem in Agda that the given
-extractor preserves the semantics.  For that one needs to relate the
-semantics of reflected terms with the semantics of the language we
-are extracting to.  Unfortunately, at the time of writing there is
-no official semantics for Agda reflected terms.  While in principle
-it is possible to come up with one manually, we believe that reflection
-system should give us the one we could actually trust.  This is our
-future work.
+Correctness of our approach thus relies on the preservation of the dynamic
+semantics by the extractor code. Ideally we would like to prove this as a
+theorem in Agda, but this is not yet possible as there is no formal semantics
+for Agda reflected terms (see future work section).
 
-## Shallow embedding and its semantics
+## Pull Requests inclusion in the paper (all reviewers)
 
-One challenge with a shallowly embedded language is how to formally define
-which terms of the host language actually form the embedding.  While we do
-not have any syntactical means to do so, as we state in line 315, any Agda
-term that is recognised by the extractor is considered to be the part of
-the embedded language.  In that sense, semantics of the embedding is always
-given by the semantics of Agda, and the language itself is defined by the
-extractor.
+Our reason for including links to PRs is because we consider them essential to
+the work presented in the paper. The important thing is not how they are
+implemented (as this will be different for each host language), but rather what
+features are required for our approach to reflection-based extraction. In
+particular, we need:
 
-## Pull Requests inclusion in the paper
+- access to the full type and typing context of a given term,
 
-We consider the changes that we made to Agda essential to the work that we
-present in the paper.  As our approach has not been tried before, obviously
-some required functionality is lacking, so we fixed it.  As for inclusion
-in the paper, unfortunately we didn't find a better way of presenting these,
-and according to our read of the definition of "lightweight double-blind"
-inclusion of the links seem to be acceptable as we do not expose our names
-or institutions, and as the work has been done specifically for this paper
-we did not have to address it in third person.
+- the ability to selectively normalise certain functions while keeping others
+untouched,
 
-We summarise the changes that we did to Agda's reflection system:
-  - allow accessing the actual typing context (telescope) of the given term;
-  - selective reduction and normalisation of the terms;
-  - access typing information of the implicit arguments after normalisation
-    -- the most challenging change.
+- access to transient terms that are erased from the syntax but can be
+reconstructed from the type.
+
+As our approach has not been tried before, some functionality was lacking so we
+had to add it. However, none of this is specific to a particular extractor (or
+even specific to extraction in general), so it only needs to be done once.
+
+According to our interpretation of the requirements of "lightweight
+double-blind", inclusion of these links seems to be acceptable: they do not
+expose our names or institutions, and as the work has been done specifically
+for this paper we did not have to address it in third person. We apologize if
+this differs from the intended interpretation.
 

### 2021-04-25 review-response.md
response part2, rev a.

@@
 
--- XXX stopped here XXX ---
+> For the author response, please respond to the next two paragraphs
+> especially.
+> 
+> Your key contribution is the idea of a "reflection-based extractor" for a
+> "shallowly-embeddable language".  What is the reflection-based extractor for
+> your Kaleidoscope example?  What are its key properties?  Can they be
+> expressed as theorems within (or possibly outside) Agda?
 
-ICFP 2021 Paper #150 Reviews and Comments
-===========================================================================
-Paper #150 Choosing is Losing: How to combine the benefits of shallow and
-deep embeddings through reflection
+Reflection based extractor for the Kaleidoscope language is define in
+`agda-extractor/Kaleid.agda` and it purpose is to translate the chosen
+subset of Agda to Kaleidoscope language.  The key properties of the translator
+is the preservation of dependent types such as Fin, `_<_`, `_>_`, 
+equivalence/inequivalence of natural numbers in the generated code via
+assertions.  Essentially, we have extended the original formulation of
+the Kaleidoscope with dependent types.  E.g. in our log2 example we ensure
+that the no dvivision by zero is possible, and we also ensure that the
+function terminates (as each Agda function has to terminate).
 
-Review #150A
-===========================================================================
+> Another key contribution is a set of changes to Agda, described in three
+> sections of the paper: 3.4, 3.5, and 4.2, each of which refer to pull
+> requests. Can you describe the changes in terms of general principles that
+> could apply in other settings?
 
-Overall merit
--------------
-C. I would not accept this paper but will not argue strongly against
-   accepting it.
+See part 1.
 
-Reviewer expertise
-------------------
-X. Expert
+> How does the performance of the code extracted from the CNN model compare to
+> the orginal?  I think the original is in APL and the extracted code is in
+> SaC.
 
-Paper summary
--------------
-1 Introduction: the idea of developing embedded programs hand-in-hand with custom
-code generators for them.
-2 Background, mostly standard stuff on Agda.  A little on quote/unquote.
-3 Basic extraction, illustrated using Kaleidoscope, a minimalist 1st order language from the LLVM team.
-Shows how to extract a Kaleidoscope program from an Agda term.
-Example of the binary logarithm.
-Use of Agda rewrite rules.
-4 Embedding the array language SaC in Agda, and extracting to it.
-A key progression compared to 3 is that SaC has a sophisticated array type system, that can be represented (more or less) within Agda.
-5 Embeds a form of APL within Agda.  The embedding is not all of APL; one key concept retained is automatic casting of scalars to vectors and arrays for use with binary operators.
+As our main goal was to demonstrate the mere possibility of embedding
+non-trivial languages, we didn't pay much attention to the performance aspect
+of the extracted code, as in some sense it is orthogonal to the process.
+However, as our code carefully follows this paper:
+https://dl.acm.org/doi/10.1145/3315454.3329960
+The paper includes the SaC version of the CNN application, and we create
+our extractor to produce the code that is as close as possible to the existing
+SaC code.  Unsurprisingly, performance of the SaC code that we generate is
+almost identical to the performance of the SaC code reported in the paper:
+it is 15 and 24 times faster for training and recognition correspondingly,
+when comparing to APL.  However, this result cannot be treated as an exhaustive
+set of measurements, as it is coming from a single example on a single machine.
 
-Pros and Cons
-+ new idea to embed DSL as a shallow embedding in Agda, but use reflection to access the syntax trees of the shallow embedding, a kind of deep embedding
-+ the shallow embedding allows properties to be type-checking, while reflection allows extraction of the verified program back to the original language
-+ three separate language examples: core language Kaleidoscope, array language SaC, and APL
-+ dependently typed array operations simply fall out of the encoding in the underlying system
-- no discussion of performance, eg, of the CNN
-- the general principles here are not expressed crisply, eg, as theorems
+> Can you say anything about what's like to program in these shallow
+> embeddings?  Are the error messages readable?  Could they be customised to
+> the embedded language?
 
-Comments for author
--------------------
-Thanks for all your work on this paper.  I learnt from reading it.
+There are two types of messages that are happening: i) Agda error messages that
+can't be specific to the embedded DSL; ii) error messages coming from the extractor.
+The latter is fully programmable, and as Agda gives basic capabilities to format
+terms and report a custom error, these can be made as precise as needed by the
+author of the extractor.
 
-Unfortunately, the main weakness to me is that there is no general theory or crisp principles here, although that may be a weakness in the presentation, rather than the underlying coding.
+> Since their deep versus shallow terminology is in the title, you should cite
+> the paper that originated the distinction between deep and shallow
+> embeddings:
+> 
+> Richard J. Boulton, Andrew D. Gordon, Michael J. C. Gordon, John Harrison,
+> John Herbert, John Van Tassel: Experience with Embedding Hardware Description
+> Languages in HOL. TPCD 1992: 129-156
 
-For the author response, please respond to the next two paragraphs especially.
-
-Your key contribution is the idea of a "reflection-based extractor" for a "shallowly-embeddable language".
-What is the reflection-based extractor for your Kaleidoscope example?
-What are its key properties?
-Can they be expressed as theorems within (or possibly outside) Agda?
-
-Another key contribution is a set of changes to Agda, described in three sections of the paper: 3.4, 3.5, and 4.2, each of which refer to pull requests.
-Can you describe the changes in terms of general principles that could apply in other settings?
-
-How does the performance of the code extracted from the CNN model compare to the orginal?  I think the original is in APL and the extracted code is in SaC.
-
-It is cool you can write APL formulas directly in Agda.
-
-Can you say anything about what's like to program in these shallow embeddings?  Are the error messages readable?  Could they be customised to the embedded language?
-
-Since their deep versus shallow terminology is in the title, you should cite the paper that originated the distinction between deep and shallow embeddings:
-
-Richard J. Boulton, Andrew D. Gordon, Michael J. C. Gordon, John Harrison, John Herbert, John Van Tassel: Experience with Embedding Hardware Description Languages in HOL. TPCD 1992: 129-156
+Of course, thank you for the reference, we are happy to include it in the paper.
 

### 2021-04-25 review-response.md
response part 2, rev b.

@@
 
+## Review #150A
+
 > For the author response, please respond to the next two paragraphs
@@
 
-Review #150B
-===========================================================================
+## Review #150B
 
-Overall merit
--------------
-C. I would not accept this paper but will not argue strongly against
-   accepting it.
+> - The introduction talks about the ability to "write a verified program
+>   hand-in-hand with an executor for it" (line 43; maybe you meant
+>   "extractor"?)),
 
-Reviewer expertise
-------------------
-Z. Some familiarity
+It seems that "executor" is adequate, as we assume to extract the code into
+a language that can compile/run the extracted piece of code.
 
-Paper summary
--------------
-This paper explores some applications of using reflection to implement code
-extraction to embedded languages.  The idea is to use reflections facilities to
-enable the translation of (subsets of) the host language (in this case, Agda)
-into an embedded language.  Such a translation might require both type and term
-translation, and might also require the generated code to include dynamic checks
-to reconcile the expressive dependend types of the host with those of the
-target.  The paper presents several case studies, including a translation of
-Agda to Kaleidoscope (a toy language included as an example from the LLVM
-framework), to a simple array language, and to APL.
+>   but this paper doesn't seem to include any semantics for the
+>   languages.
 
-Comments for author
--------------------
-# Review
+See part 1: embedded language is a subset of Agda, therefore it shares Agda's
+semantics.
 
-I had a very hard time getting anything out of this paper.  The authors espouse
-the benefits of combining "shallow" and "deep" embeddings, ostensibly with the
-purposes of doing things like verified extraction or code transformations, but
-the paper doesn't really deliver on that promise.  Moreover, I found the
-presentation very confusing.  I cannot support this paper for acceptance.
+>   The "verification results" seem to be very limited in nature --
+>   providing, e.g., some additional guarantees on top of APL's weak shape-based
+>   type system.
 
-# Comments
+This is more "correct-by-construction programming" rather than classical verification.
+As dependent types can encode virtually any constraints on terms, the type system
+guarantees that these hold statically, and extractor preserves (modulo extractor
+correctness) these constraints in the generated code, we can talk about a verified
+program.
 
-- The introduction talks about the ability to "write a verified program
-  hand-in-hand with an executor for it" (line 43; maybe you meant
-  "extractor"?)), but this paper doesn't seem to include any semantics for the
-  languages.  The "verification results" seem to be very limited in nature --
-  providing, e.g., some additional guarantees on top of APL's weak shape-based
-  type system.  Moreover, since assertations that capture those stronger
-  constraints are propagated into the output, I would expect there to be some
-  kind of theorem stating that those assertaions are never triggered (under
-  reasonable assumptions).  There aren't (as far as I can see), and explicitly
-  stated correctness or verification theorems mentioned in the paper. Is
-  there  some kind of correctness guarantee that you get when translating from
-  APL to SaC via your infrastructure?
+>   Moreover, since assertations that capture those stronger
+>   constraints are propagated into the output, I would expect there to be some
+>   kind of theorem stating that those assertaions are never triggered (under
+>   reasonable assumptions).  There aren't (as far as I can see), and explicitly
+>   stated correctness or verification theorems mentioned in the paper. Is
+>   there  some kind of correctness guarantee that you get when translating from
+>   APL to SaC via your infrastructure?
 
-- I'm not sure that I would consider the encodings of Kaleidoscope or SaC in
-  this paper as "shallow embeddings": my understanding is that a shallow
-  embedding uses the features of the host language (in this case, Agda) to
-  implement the _semantics_ of the embedded language (often using host-level
-  functions to represent embedding-level variable binding, etc.).  However, I
-  didn't see any semantics ascribed to Kaleidoscope, and only minimally so for
-  SaC.  It looks like this code doesn't do much more than provide the abstract
-  syntax for the embedded languages in question -- based on the definitions in
-  this paper could you prove anything about the behavior of a program extracted
-  to Kaleidoscope?  The APL embedding seems like a true embedded language in the
-  sense that its semantics are implemened in Agda.
+See Part 1.
 
-- In several places, the paper mentions limitations of Agda's reflection API
-  that were changed or enhanced while doing this work (e.g. paragraphs near line
-  332, 362, 829), but these are mentioned mostly in passing and not described in
-  enough detail to be meanginful in any way to the reader.  Pointing to github
-  pull requests is not appropriate for a scholarly paper (and isn't kosher for
-  double-blind reviewing).  These anecdotes made me wonder what the point of
-  this paper is, since it's apparently *not* intended to explain any of the
-  technical details involved with these changes.
+> - I'm not sure that I would consider the encodings of Kaleidoscope or SaC in
+>   this paper as "shallow embeddings": my understanding is that a shallow
+>   embedding uses the features of the host language (in this case, Agda) to
+>   implement the _semantics_ of the embedded language (often using host-level
+>   functions to represent embedding-level variable binding, etc.).  However, I
+>   didn't see any semantics ascribed to Kaleidoscope, and only minimally so for
+>   SaC.  It looks like this code doesn't do much more than provide the abstract
+>   syntax for the embedded languages in question -- based on the definitions in
+>   this paper could you prove anything about the behavior of a program extracted
+>   to Kaleidoscope?  The APL embedding seems like a true embedded language in the
+>   sense that its semantics are implemened in Agda.
 
-- The translation of reflected terms into Kaleidoscope doesn't seem to be that
-  complicated, but I would have appreciated if the discussion starting at line
-  557 had walked though the code of `kompile-term` a bit more explicitly.
+Additionally to explanations in Part 1, all the three languages share Agda's semantics.
+However, it happened so, that many Kaleidoscope and SaC constructs, such as Nat,
+arithmetic expressions, functions, etc. are also built-in Agda's constructs.
+Therefore, there is no obvious Agda function that implements some embedded structure.
+However, reflection makes the entire Agda AST available, therefore as long as
+Agda terms have semantics, so does the embedding.
 
-# Details
+> - In several places, the paper mentions limitations of Agda's reflection API
+>   that were changed or enhanced while doing this work (e.g. paragraphs near line
+>   332, 362, 829), but these are mentioned mostly in passing and not described in
+>   enough detail to be meanginful in any way to the reader.  Pointing to github
+>   pull requests is not appropriate for a scholarly paper (and isn't kosher for
+>   double-blind reviewing).  These anecdotes made me wonder what the point of
+>   this paper is, since it's apparently *not* intended to explain any of the
+>   technical details involved with these changes.
 
-- line 184: says that the primitives are `quote` and `unquote` and that they
-  operate "as follows", but the subsequent code uses `macro` and `quote` has
-  already been explained above.
+See Part 1.
 
-- line 331: "contains any about strings" $\to$ "contains any strings"
+> - The translation of reflected terms into Kaleidoscope doesn't seem to be that
+>   complicated, but I would have appreciated if the discussion starting at line
+>   557 had walked though the code of `kompile-term` a bit more explicitly.
 
-- line 355: This definition of decidable equality is unintelligible to a reader
-  with no Agda experience.  (Also, it might be good to mention earlier that it's
-  common for Agda identifiers to include symbols that are usualy considered
-  reserved).
+We are happy to adjust this bit of the paper.
 
-- line 787:
+> - line 355: This definition of decidable equality is unintelligible to a reader
+>   with no Agda experience.  (Also, it might be good to mention earlier that it's
+>   common for Agda identifiers to include symbols that are usualy considered
+>   reserved).
 
-- line 789: The "lx type is a type of valid indices..." should probably be "lx d s type is..."
+Happy to expand on this as well.
+
+> - line 789: The "lx type is a type of valid indices..." should probably be "lx d s type is..."
+
+You are right, however it is common to use "type" and "type family" interchangeably.
+We are happy to clarify this.
 
 - line 1208: Seems pretty late in the paper to bury the lede about not *really* doing any kind of verification.
+
+This is because our main focus was on providing the framework in which it is
+possible to define such embeddings in the first place.  When we extract Ocaml
+from Coq it is as unverified as what we propose.
 

### 2021-04-25 review-response.md
Part 2, reviewer C.

@@
 
-Review #150C
-===========================================================================
+## Review #150C
 
-Overall merit
--------------
-A. I will champion accepting this paper.
+> What I'm less convinced by
+> is the ease of use for the _implementor_. In particular, promises of the introduction
+> like "No need to touch internals of the host language" are true only in spirit. Yes,
+> in a language with ideal reflection capabilities what the authors describe would be
+> even more convenient than described in the paper, but even for this work the authors
+> had to submit multiple pull requests to change Agda internals and expose additional
+> capabilities. A deep embedding for Kaleidoscope for example sounds simple enough to be
+> much more natural to implement - but I guess then you wouldn't enjoy the ease-of-use
+> that we're seeing here.
 
-Reviewer expertise
-------------------
-Y. Knowledgeable
+Yes, you are right.  However, as we are trying a novel approach that hasn't been tried
+before (to our knowledge), we don't expect all the necessary API to be present.
+As you can see from Part 1, our extensions are not language-specific.  They are
+of a general nature, and given a few more language embeddings will be tried out,
+the API should stabilise rather quickly.
 
-Paper summary
--------------
-This paper attempts to answer the choice between shallow and deep embeddings
-(in particular in a dependently typed context), by refusing to answer the question
-in the first place. Instead, the authors rely on compile-time metaprogramming facilities
-in a host language to specify a subset of the host language (in this case Agda) that
-serves as the shallow-ish embedding that can be extracted to the target language of choice,
-while still allowing for reasoning about its code using all the convenience of a
-language like Agda. The paper then proceeds to give two extensive and convincing case
-studies using this approach.
+> - L. 27: Not the canonical citation for Coq I think. Usually people reach for
+>   the reference manual
 
-Comments for author
--------------------
-This paper was _really_ fun to read. It opens with an intriguing idea - can we use
-reflection facilities in a dependently typed language - to get the best out of both
-worlds: those of shallow and deep embeddings. And it technically delivers that!
-The 2/2.5 case studies presented demonstrate convincingly that this is a viable
-approach - the matrix multiplication example in particular is an excellent
-demonstration that _users_ of the embedding can enjoy strong static guarantees
-in a seamless manner like in a shallow embedding. What I'm less convinced by
-is the ease of use for the _implementor_. In particular, promises of the introduction
-like "No need to touch internals of the host language" are true only in spirit. Yes,
-in a language with ideal reflection capabilities what the authors describe would be
-even more convenient than described in the paper, but even for this work the authors
-had to submit multiple pull requests to change Agda internals and expose additional
-capabilities. A deep embedding for Kaleidoscope for example sounds simple enough to be
-much more natural to implement - but I guess then you wouldn't enjoy the ease-of-use
-that we're seeing here.
+Thanks, happy to adjust that.
 
-Still, even though the paper reads like a tour-de-force in Agda-fu, it is really well written,
-with multiple carefully selected and well explained examples. Do I think that after
-reading this paper I will be reaching when implementing my next DSL for this technique
-(or even that I'm able to reach for this technique without significant investment in
-metaprogramming internals)? Perhaps; probably not. But it is a novel first step taking
-an old dilemma in a very interesting direction, while showing concrete potential.
+> - Pull request links should probably have been elided for the review version
+> (I haven't clicked them yet as of the time of this writing, but I'm assuming
+> the break double blind)
 
-======== Minor ==========
-
-- L. 27: Not the canonical citation for Coq I think. Usually people reach for the reference manual
-
-- Pull request links should probably have been elided for the review version
-(I haven't clicked them yet as of the time of this writing, but I'm assuming the break double blind)
+See Part 1.

### 2021-04-25 review-response.md
polishing response

@@
 without implementing a full typechecker, as you share one with Agda. Extraction
-allows us to translate our DSL to a target language for further execution. We
+allows us to translate our DSL to a target language for further execution.  This
+translation is expressed within Agda without necessity to modify the compiler.  We
 expect two possible scenarios: 1) translating an existing (dependently typed)
@@
 guarantees that none of the inserted assertions will be triggered at run-time.
-When only part of a program is extracted from Agda, assertions can only fire at
-the boundaries with unsafe code.
+When only part of a program is extracted from Agda (which is perfectly possible),
+assertions can only fire at the boundaries with unsafe code.
 
@@
 
-Reflection based extractor for the Kaleidoscope language is define in
+Reflection based extractor for the Kaleidoscope language is defined in
 `agda-extractor/Kaleid.agda` and it purpose is to translate the chosen
@@
 assertions.  Essentially, we have extended the original formulation of
-the Kaleidoscope with dependent types.  E.g. in our log2 example we ensure
-that the no dvivision by zero is possible, and we also ensure that the
+the Kaleidoscope to include dependent types.  E.g. in our log2 example we ensure
+that the dvivision by zero is impossible, and we also ensure that the
 function terminates (as each Agda function has to terminate).
@@
 As our main goal was to demonstrate the mere possibility of embedding
-non-trivial languages, we didn't pay much attention to the performance aspect
-of the extracted code, as in some sense it is orthogonal to the process.
-However, as our code carefully follows this paper:
+non-trivial languages, we didn't focus too much on the performance aspect
+of the extracted code in the text of the paper.  In some sense it is orthogonal
+to the extraction process.  However, our code carefully follows this paper:
 https://dl.acm.org/doi/10.1145/3315454.3329960
-The paper includes the SaC version of the CNN application, and we create
+The paper includes the SaC version of the CNN application, and we created
 our extractor to produce the code that is as close as possible to the existing
@@
 it is 15 and 24 times faster for training and recognition correspondingly,
-when comparing to APL.  However, this result cannot be treated as an exhaustive
+when comparing to APL (recall that APL runs in an interpreter, but SaC is
+a compiled language).  However, this result cannot be treated as an exhaustive
 set of measurements, as it is coming from a single example on a single machine.
@@
 
-There are two types of messages that are happening: i) Agda error messages that
+There are two types of error messages that can be triggered: i) Agda error messages that
 can't be specific to the embedded DSL; ii) error messages coming from the extractor.
-The latter is fully programmable, and as Agda gives basic capabilities to format
+The latter is fully programmable; and as Agda gives basic capabilities to format
 terms and report a custom error, these can be made as precise as needed by the
-author of the extractor.
+author of the extractor.  Practically, as long as functions are reasonably sized
+(as in our case) the error messages are readable.
 
@@
 This is more "correct-by-construction programming" rather than classical verification.
-As dependent types can encode virtually any constraints on terms, the type system
+As dependent types can encode virtually any constraints on terms, and the type system
 guarantees that these hold statically, and extractor preserves (modulo extractor
@@
 arithmetic expressions, functions, etc. are also built-in Agda's constructs.
-Therefore, there is no obvious Agda function that implements some embedded structure.
+Therefore, there is no obvious Agda function that implements some embedded structures.
 However, reflection makes the entire Agda AST available, therefore as long as

### 2021-04-25 review-response.md
more polishing.

@@
 assertions.  Essentially, we have extended the original formulation of
-the Kaleidoscope to include dependent types.  E.g. in our log2 example we ensure
-that the dvivision by zero is impossible, and we also ensure that the
+the Kaleidoscope to include dependent types.  E.g. in our log2 example types ensure
+that the dvivision by zero is impossible, and we know that the extracted
 function terminates (as each Agda function has to terminate).
@@
 https://dl.acm.org/doi/10.1145/3315454.3329960
-The paper includes the SaC version of the CNN application, and we created
+The paper includes a SaC version of the CNN application, and we created
 our extractor to produce the code that is as close as possible to the existing
@@
 it is 15 and 24 times faster for training and recognition correspondingly,
-when comparing to APL (recall that APL runs in an interpreter, but SaC is
-a compiled language).  However, this result cannot be treated as an exhaustive
+when comparing to APL (recall that APL is an interpreter, but SaC is
+a compiled language).  However, this result should not be treated as an exhaustive
 set of measurements, as it is coming from a single example on a single machine.
@@
 
-There are two types of error messages that can be triggered: i) Agda error messages that
-can't be specific to the embedded DSL; ii) error messages coming from the extractor.
-The latter is fully programmable; and as Agda gives basic capabilities to format
-terms and report a custom error, these can be made as precise as needed by the
+There are two types of error messages that can be triggered: i) Agda error messages;
+ii) error messages coming from the extractor.  The former can't be specific to the
+embedded DSL.  The latter is fully programmable; and as Agda includes basic capabilities
+to format terms and report custom errors, these can be made as precise as needed by the
 author of the extractor.  Practically, as long as functions are reasonably sized
@@
 
-We are happy to adjust this bit of the paper.
+We will adjust this bit of the paper.
 
@@
 
-Happy to expand on this as well.
+We are happy to expand on this.
 
@@
 
-You are right, however it is common to use "type" and "type family" interchangeably.
-We are happy to clarify this.
+You are right. We used "type" and "type family" interchangeably.  We will clarify this.
 
@@
 
-This is because our main focus was on providing the framework in which it is
+This is because our main focus is to provide the framework in which it is
 possible to define such embeddings in the first place.  When we extract Ocaml
